---
title: "Greedy Algorithm"
author: "Luis David Bedon Gomez"
date: "5 9 2017"
header-includes:
    - \usepackage{hyperref}
output:
  pdf_document: default
  html_document: default
---

The Greedy Algorithm is an optimization algorithm, that picks up the objects with the greatest value until the constraint is reached. See e.g.  
[Data Science MIT](https://ocw.mit.edu/courses/electrical-engineering-and-computer-science/6-0002-introduction-to-computational-thinking-and-data-science-fall-2016/lecture-slides-and-files/)

If there are more objects with a subsequent value and the value of the objects chosen so far is less then the constraint, the objects with the now highest value are chosen and so on.


# Building a greedy algorithm:

## Generate/import file with the objects and values to optimize for given a constraint:


```{r generation}
# Generate the data
set.seed(120)
nobjects<-15
table1<-data.frame(object=letters[1:nobjects],
                   number=rpois(nobjects,3),
                   value=round(1+abs(rnorm(nobjects,mean=nobjects*.6,sd=nobjects*.6)),0))

head(table1,8)
```


# Greedy-Algorithm


## Algorithm

The Greedy-Algorithm should:

- Calculate the global total value
- Arrange by total value
- Begin the computation

This can be achieved as follows:

### 1. Reorganize objects by total value in descent order:
```{r,message=FALSE}
library(dplyr)
table2<-table1 %>% mutate(total.value=number*value) %>% arrange(desc(total.value))
print(table2)
```

Note, that by creating the new column *total.value* and sorting it from high to low, the order changes in respect to the value for each object!! 

### 2. Set a constrain:
  
```{r}
constraint<-sum(table2$total.value)/1.1
print(constraint)
```

### 3. Run the iteration:

```{r}
k<-1
result<-{}
resvalue<-0
for(i in 1:length(table2$object)){
  if(resvalue<=constraint){
    for(j in 1:table2$number[i]) {
       if((resvalue+table2[i,3])<=constraint) {
         resvalue<-resvalue+table2[i,3]
         result[k]<-as.character(table2$object[i])
         k<-k+1
       }
    } 
  } 
}
print(resvalue)
print(result)
print(paste("Constraint=",constraint,"vs. Reached Value=",resvalue))


```
## Visualization

```{r,message=FALSE,echo=FALSE}
library(ggplot2)
library(extrafont)
#font_import(pattern="Interstate-Regular")

# Cumulative table with all objects and values
numobject<-function(x) as.character(rep(table1$object[x],table1$num[x]))
cumobject<-sapply(1:length(table1$object),numobject)
table3<-data.frame(object=unlist(cumobject))
table3$value<-sapply(1:length(table3$object),
                     function(x) table1$value[table1$object==table3$object[x]])

subset_t3<-function(x) table3[table3$object==as.character(x),2]
cumsubset_t3<-function(y) {
              sapply(1:length(subset_t3(y)),
                     function(x) sum(subset_t3(y)[1:x]))
              }
cumallsubset<-sapply(as.character((table1$object)),cumsubset_t3)
table3$cum<-unlist(cumallsubset)

# Cumulative table after algorithm results
s<-(sapply(result,function(x) grep(x,table2$object)))

table4<-data.frame(object=table2$object[s],value=table2$value[s])
table4$cum<-sapply(1:(length(table4$value)),function(x) sum(table4$value[1:x]))

lresult<-length(table(result))

# Plot with all objects and values
ggplot()+
  geom_tile(aes(x=table3$object,
                y=table3$cum-table3$value/2,fill=table3$value),
            width=.4,
            height=table3$value,
            colour="white")+
  geom_hline(yintercept = constraint,colour=rgb(.95,.55,.55))+
  annotate("text",x=2,y=constraint-10,label= "Constraint",colour=rgb(.95,.55,.55))+
  scale_fill_gradient2(low=rgb(.55,.45,.65,1),
                       high=rgb(.35,.65,.25,.8),
                       midpoint = 25,mid=rgb(.85,.85,.55,.2))+
  ggtitle(label="Hypothetical Objects with Number and Values\nsorted alphabetically")+
  xlab("Hypothetical Objects")+
  ylab("Hypothetical Value")#+
  #theme(text=element_text(family="Interstate"))

# Plot showing the algorithm resuts for the given constraint
ggplot()+
  geom_tile(aes(x=reorder(table4$object,table4$cum),
                y=table4$cum-table4$value/2,fill=table4$value),
            width=.4,
            height=table4$value,
            colour="white")+
  geom_hline(yintercept = constraint,colour=rgb(.95,.55,.55))+
  annotate("text",x=2,y=constraint-10,label= "Constraint",colour=rgb(.95,.55,.55))+
  scale_fill_gradient2(low=rgb(.55,.45,.65,1),
                       high=rgb(.35,.65,.25,.8),
                       midpoint = 25,mid=rgb(.85,.85,.55,.2))+
  ggtitle(label="Greedy Algorithm choosing the highest\npossible value of objects")+
  xlab("Hypothetical Objects")+
  ylab("Hypothetical Value")
  


  

```



## Some considerations:

- The greedy algorithm should fail by maximal the modulo of the highest possible value to the respective highest value available.

- The greedy algorithm resambles a decomposition in prime numbers

- It represents a up-down algorithm with no branches

- Possible uses could be:

      - Choosing coins and notes to give the change back after paying cash
  
  
  
  
## Appendix

The code for the plots:
```{r,message=FALSE,echo=TRUE,results="hide",fig.show='hide'}
library(ggplot2)
library(extrafont)
#font_import(pattern="Interstate-Regular")

# Cumulative table with all objects and values
numobject<-function(x) as.character(rep(table1$object[x],table1$num[x]))
cumobject<-sapply(1:length(table1$object),numobject)
table3<-data.frame(object=unlist(cumobject))
table3$value<-sapply(1:length(table3$object),
                     function(x) table1$value[table1$object==table3$object[x]])

subset_t3<-function(x) table3[table3$object==as.character(x),2]
cumsubset_t3<-function(y) {
              sapply(1:length(subset_t3(y)),
                     function(x) sum(subset_t3(y)[1:x]))
              }
cumallsubset<-sapply(as.character((table1$object)),cumsubset_t3)
table3$cum<-unlist(cumallsubset)

# Cumulative table after algorithm results
s<-(sapply(result,function(x) grep(x,table2$object)))

table4<-data.frame(object=table2$object[s],value=table2$value[s])
table4$cum<-sapply(1:(length(table4$value)),function(x) sum(table4$value[1:x]))

lresult<-length(table(result))

# Plot with all objects and values
ggplot()+
  geom_tile(aes(x=table3$object,
                y=table3$cum-table3$value/2,fill=table3$value),
            width=.4,
            height=table3$value,
            colour="white")+
  geom_hline(yintercept = constraint,colour=rgb(.95,.55,.55))+
  annotate("text",x=2,y=constraint-10,label= "Constraint",colour=rgb(.95,.55,.55))+
  scale_fill_gradient2(low=rgb(.55,.45,.65,1),
                       high=rgb(.35,.65,.25,.8),
                       midpoint = 25,mid=rgb(.85,.85,.55,.2))+
  ggtitle(label="Hypothetical Objects with Number and Values\nsorted alphabetically")+
  xlab("Hypothetical Objects")+
  ylab("Hypothetical Value")#+
  #theme(text=element_text(family="Interstate"))

# Plot showing the algorithm resuts for the given constraint
ggplot()+
  geom_tile(aes(x=reorder(table4$object,table4$cum),
                y=table4$cum-table4$value/2,fill=table4$value),
            width=.4,
            height=table4$value,
            colour="white")+
  geom_hline(yintercept = constraint,colour=rgb(.95,.55,.55))+
  annotate("text",x=2,y=constraint-10,label= "Constraint",colour=rgb(.95,.55,.55))+
  scale_fill_gradient2(low=rgb(.55,.45,.65,1),
                       high=rgb(.35,.65,.25,.8),
                       midpoint = 25,mid=rgb(.85,.85,.55,.2))+
  ggtitle(label="Greedy Algorithm choosing the highest\npossible value of objects")+
  xlab("Hypothetical Objects")+
  ylab("Hypothetical Value")
  


  

```
